---
title: "README"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `flowUtilities`

[Report a bug](https://github.com/stemcellbioengineering/flowUtilities/issues)

`flowUtilities` is an R package that simplifies flow cytometry data analysis by providing convenient wrapper functions and utilities for common tasks in the flow cytometry analysis workflow. The package builds on popular libraries within the [Bioconductor](https://bioconductor.org/) ecosystem to streamline data import, gating, visualization, and statistical analysis.

## Installation

After installing all dependencies from Bioconductor, the `flowUtilities` package can be installed from GitHub:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("stemcellbioengineering/flowUtilities")
```

## Dependencies

`flowUtilities` requires the following Bioconductor packages:

```{r eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("flowCore", "flowWorkspace", "flowStats", "ggcyto"))
```

## Documentation

For a workflow example, see the package vignette:

```{r eval=FALSE}
browseVignettes("flowUtilities")
```

## Contributions

Contributors are welcome! If you have an idea for a new feature you can [issue a pull request](https://github.com/stemcellbioengineering/flowUtilities/pulls).

## License

This project is licensed under the GNU General Public License v3.0 - see the LICENSE file for details.
